A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2021-05-05 submission/manuscript.rmd
minor changes in discussion

In recent years several methods for the evaluation of treatment effect
heterogeneity have been developed in the setting of RCTs [@Rekkas2020].
However, low power and restricted prior knowledge on the mechanisms of variation
in treatment effect are often inherent in RCTs, which are usually adequately
powered only for the analysis of the primary outcome. Observational databases
contain a large amount of information on treatment assignment and outcomes of
interest, while also capturing key patient characteristics. Our framework
provides a standardized approach that can be used to leverage available
information from these data sources, allowing for large-scale risk-based
assessment of treatment effect heterogeneity. Multiple outcomes can be evaluated
in patient subgroups of similar baseline outcome risk. Multiple outcome risk
stratification schemes can also be considered. However, this should be done with
caution, as it may hinder the interpretability of the results, in a similar
manner as typical subgroup analyses. The standardized nature of the framework
enables transportability to multiple databases, provided that they are mapped to
the OMOP-CDM.

Recently, guidelines on the application of risk modeling approaches for the
assessment of heterogeneity of treatment effect in RCT settings have been
proposed [@Kent2019; @PathEnE]. Our framework aims to translate these
guidelines to the observational setting while also providing a toolset for its
implementation. Several considerations need to be made. First, estimates may be
biased due to the observational nature of the data. We attempt to account for
potential confounding by estimating propensity scores within strata of predicted
risk. These scores are estimated using regularized logistic regression on a
large set of pre-defined covariates. However, such approaches do not account for
unobserved confounding [@Liu2013]. Several sensitivity analyses have been
proposed in the literature for measuring the robustness of results in the
presence of unobserved confounding. Another approach is to calibrate estimates
and confidence intervals based on a large set of negative controls
[@Schuemie2014; @Schuemie2571]. Negative controls are treatment-outcome pairs
for which a null effect has been established. Estimating these effects within
available data provides an approximation of the null distribution that can be
used to empirically recalibrate effect estimates. Future work may extend our
framework with this type of analyses.

@@

Our contribution is a translation of the PATH statement principles to the OHDSI
methods library [@Kent2019; @PathEnE]. Our method encourages open science as
it requires accurate definition of the research questions translated into clear
and reproducible cohort definitions that can easily be shared among researchers.
Our R-package provides a standardized stepwise procedure for the assessment of
HTE. This enables source code to be easily shared and evaluated, allowing for
reproducible research. Researchers with access to different databases mapped to
OMOP-CDM can also very easily extend their overall analyses with risk-based
assessment of treatment effect heterogeneity. This enables collaboration among
multiple sites with access to different patient populations. We propose that the
framework is implemented any time treatment effect estimation in
high-dimensional observational data is undertaken.

>>>>

In recent years several methods for the evaluation of treatment effect
heterogeneity have been developed in the setting of RCTs [@Rekkas2020].
However, low power and restricted prior knowledge on the mechanisms of variation
in treatment effect are often inherent in RCTs, which are usually adequately
powered only for the analysis of the primary outcome. Observational databases
contain a large amount of information on treatment assignment and outcomes of
interest, while also capturing key patient characteristics. Our framework
provides a standardized approach that can be used to leverage available
information from these data sources, allowing for large-scale risk-based
assessment of treatment effect heterogeneity. Multiple outcomes can be evaluated
in patient subgroups of similar baseline outcome risk. Multiple outcome risk
stratification schemes can also be considered. However, this should be done with
caution, as it may hinder the interpretability of the results, in a similar
manner as typical subgroup analyses.

Recently, guidelines on the application of risk modeling approaches for the
assessment of heterogeneity of treatment effect in RCT settings have been
proposed [@Kent2019; @PathEnE]. Our framework aims to translate these
guidelines to the observational setting while also providing a toolkit for its
implementation within OHDSI. It encourages open science as it requires accurate
definition of the research questions translated into clear and reproducible
cohort definitions that can easily be shared among researchers. Researchers with
access to different databases mapped to OMOP-CDM can also very easily extend
their overall analyses with risk-based assessment of treatment effect
heterogeneity. This enables collaboration among multiple sites with access to
different patient populations. We propose that the framework is implemented any
time treatment effect estimation in high-dimensional observational data is
undertaken.

Several considerations need to be made. First, estimates may be
biased due to the observational nature of the data. We attempt to account for
potential confounding by estimating propensity scores within strata of predicted
risk. These scores are estimated using regularized logistic regression on a
large set of pre-defined covariates. However, such approaches do not account for
unobserved confounding [@Liu2013]. Several sensitivity analyses have been
proposed in the literature for measuring the robustness of results in the
presence of unobserved confounding. Another approach is to calibrate estimates
and confidence intervals based on a large set of negative controls
[@Schuemie2014; @Schuemie2571]. Negative controls are treatment-outcome pairs
for which a null effect has been established. Estimating these effects within
available data provides an approximation of the null distribution that can be
used to empirically recalibrate effect estimates. Future work may extend our
framework with this type of analyses.

@@
